---
title: "Differentials of moving frames"
author: "Li,Zhi"
date: "9/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Transporting Vectors

Looking at the graph.We have a black circle, it is the $\xi'$ circle. Because the $\xi=\{x,y\}$ is for Cartesian coordinates, we use $\xi'=\{r,\theta\}$ for polar coordinates, we add a $'$ just to say polar. When we want to plot polar coordinates, we need to convert it to Cartesian coordinates by $$\begin{align} x &= r\,\cos\theta \\ y &= r\,\sin\theta \end{align}$$ 
```{r code1}

theta=seq(0,2*pi,by=.1*pi)
r=5
x=r*cos(theta)
y=r*sin(theta)

plot(x,y,ylim=c(-5,20),xlim=c(-5,40))

points(x=3,y=4,col="red")
points(x=4,y=5,col="red")

points(x=4,y=3,col="red")
points(x=5,y=4,col="red")

r=5;theta=0.9272952# origin 3,4
basis=matrix(c(cos(theta),-(r * sin(theta)),sin(theta),r * cos(theta)) ,nrow=2,byrow=1)

ang=seq(0,2*pi,by=.1*pi)
n=length(ang)
data1=rbind(rep(0,n),ang)
thetaaxis=apply(data1,2,function(x) basis%*%x)
points(x=(thetaaxis[1,]+3),y=(thetaaxis[2,]+4),col="blue",type="p",pch=3)

ang=rep(0,n)
data2=rbind((1:n-1),ang)
raxis=apply(data2,2,function(x) basis%*%x)
points(x=(raxis[1,]+3),y=(raxis[2,]+4),col="blue",type="p",pch=4)

r=5;theta=0.6435011# origin 4,3
basis=matrix(c(cos(theta),-(r * sin(theta)),sin(theta),r * cos(theta)) ,nrow=2,byrow=1)

ang=seq(0,2*pi,by=.1*pi)
n=length(ang)
data3=rbind(rep(0,n),ang)
thetaaxis=apply(data3,2,function(x) basis%*%x)
points(x=(thetaaxis[1,]+3),y=(thetaaxis[2,]+4),col="blue",type="p",pch=3)

ang=rep(0,n)
data4=rbind((1:n-1),ang)
raxis=apply(data4,2,function(x) basis%*%x)
points(x=(raxis[1,]+4),y=(raxis[2,]+3),col="blue",type="p",pch=4)
```

There is a vector $V=\{x=1,y=1\}$ in red at starting point $\{3,4\}$ is transported to a new starting point $\{4,3\}$. The transportation does not make changes on vector $V=\{1,1\}$, it just places $V$ onto a new starting point.

## The tangent basis

The tangent basis are in blue. From our differential geometry we knew tangent basis are partial derivatives go like this $$\begin{align} x &= r\,\cos\theta \\ y &= r\,\sin\theta \end{align}$$



$$\hat e_i=\left[\begin{matrix} \frac{\partial x}{\partial r}& \frac{\partial x}{\partial \theta}\\ \frac{\partial y}{\partial r}& \frac{\partial y}{\partial \theta} \end{matrix}\right]=\left[\begin{matrix} \cos(\theta) & -r\sin(\theta)\\\sin(\theta)  &r\cos(\theta)  \end{matrix}\right]$$


```{r code2}
require(Deriv)
xi=c("r*cos(theta)","r*sin(theta)")
#partial derivative basis vectors
basis1=sapply(xi,function(m) Deriv(m,"r"))
basis2=sapply(xi,function(m) Deriv(m,"theta"))
cbind(basis1,basis2) #contravariant matrix
```

Say we have a point $\{V^r,V^{\theta}\}$ in the tangent space. We want to plot it onto the graph by Cartesian coordinates $\{V^x,V^y\}$. We can use the tangent basis transformation $$\left[\begin{matrix}V^x\\V^{y}\end{matrix}\right]=\left[\begin{matrix} \frac{\partial x}{\partial r}& \frac{\partial x}{\partial \theta}\\ \frac{\partial y}{\partial r}& \frac{\partial y}{\partial \theta} \end{matrix}\right](\xi')\left[\begin{matrix}V^r\\V^{\theta}\end{matrix}\right]=\left[\begin{matrix} \cos(\theta) & -r\sin(\theta)\\\sin(\theta)  &r\cos(\theta)  \end{matrix}\right](\xi')\left[\begin{matrix}V^r\\V^{\theta}\end{matrix}\right]$$The $(\xi')$ notation means we do have to specify the tangent basis's origin in term of $r,\theta$, most of the texts would drop the $(\xi')$ notation to show matrix transformation. 

## Inside tangent space 

Say we want to transform the vector $\{1,1\}$ into the tangent space, again we do have to specify the origin of tangent basis in term of $x,y$.  We convert Cartesian coordinates into polar coordinates $$r=\sqrt{x^2+y^2}\\\theta=\tan^{-1}(\frac{y}{x})$$ $$\left[\begin{matrix}V^r\\V^{\theta}\end{matrix}\right]=\left[\begin{matrix} \frac{\partial r}{\partial x}& \frac{\partial r}{\partial y}\\ \frac{\partial \theta}{\partial x}& \frac{\partial\theta}{\partial y} \end{matrix}\right](\xi)\left[\begin{matrix}V^x\\V^{y}\end{matrix}\right]$$ 
```{r code3}
x=3
y=4
sqrt(x^2+y^2)
atan(y/x)

x=4
y=3
sqrt(x^2+y^2)
atan(y/x)

xi.=c("sqrt(x^2+y^2)","atan(y/x)")
V=c(1,1)

#partial derivative basis vectors
sapply(xi.,function(m) Deriv(m,"x"))
sapply(xi.,function(m) Deriv(m,"y"))

x=3;y=4
basis1=c(x/sqrt(x^2 + y^2),-(y/(x^2 * ((y/x)^2 + 1))))#copy pasted results
basis2=c(y/sqrt(x^2 + y^2),1/(x * ((y/x)^2 + 1)))
cbind(basis1,basis2)%*%V

x=4;y=3
basis1=c(x/sqrt(x^2 + y^2),-(y/(x^2 * ((y/x)^2 + 1))))
basis2=c(y/sqrt(x^2 + y^2),1/(x * ((y/x)^2 + 1)))
cbind(basis1,basis2)%*%V
```
We see inside the 2 tangent spaces we have built, the vector $\{1,1\}$ has values $\{1.4,-0.04\}$ and $\{1.4,0.04\}$, which matches our results have showed in the graph above. 

## The 2 differentials

Here I only state some results without proof, the proofs are not hard only tedious. The tangent basis are moving, so they are called the moving frames. The vector $V$ in red is also moving. The 2 differentials are $$d\hat e_i=\omega^i{}_{j} \hat e_i $$ $$dV'^i=\Gamma'^i{}_{jk}d\xi'^jV'^k$$ These 2 formulas are the same thing. Note the $V$ used above conveniently became $V'$ in the 2nd formula, the proof tells why.

Some important facts$$\begin{equation} \omega_{ij} = g_{ik} \,\omega^k{}_j \end{equation}$$ $$\displaystyle{ \Gamma_{kij}
  = {\Gamma^m}_{ij}g_{mk}
  = \frac{\partial \mathbf{e}_{i}}{\partial x^j} \cdot \mathbf{e}^{m} g_{mk}
  = \frac{\partial \mathbf{e}_{i}}{\partial x^j} \cdot \mathbf{e}_{k}
 }$$ $$\begin{equation} \omega_{ij}  = \Gamma_{ijk}\,\sigma^k   \end{equation}$$
 
We can have $$\omega_{ij} = g_{ik} \,\omega^k{}_j =\Gamma_{kij}\sigma^k
  = {\Gamma^m}_{ij}g_{mk}\sigma^k$$ Cancel $g_{ij}$ we have $$\omega^i{}_{j}=\Gamma^i{}_{jk}\sigma^k$$
I got $\sigma^k$ notation from another book, it is equivalent to $d\xi'$. We see the other book also drops the $'$. We say $\omega^i{}_j$ and $\omega_{ij}$ are the connections.

