---
title: "Differential Forms 1"
author: "Li,Zhi"
date: "7/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## One-Forms 

A 1-form is a differential expression looks like:
$$Pdx+Qdy$$
where $P=P(x,y)$ and $Q=Q(x,y)$ are functions of $x,y$. In the textbooks, the definition does require functions to be $C^1$ or $C^2$, $C$ means continuous with first/second derivative.

From our elementary calculus we had this formula
$$d(uv)=udv+vdu$$
More generally, if $f(x,y)$ is $C^1$, then
$$df=\frac{\partial f}{\partial x}dx+\frac{\partial f}{\partial y}dy$$
This is actually called total differential or exterior derivative or the Master Formula in vector calculus. 


Similarly with $f(x,y,z)$ we have
$$df=\frac{\partial f}{\partial x}dx+\frac{\partial f}{\partial y}dy+\frac{\partial f}{\partial z}dz$$
The idea is that differential of 0-form is a 1-form. 

Variables such as x,y,z are 0-forms. Functions contain $dx, dy, dz$, are called 1-forms. The $dx\wedge dy, dy\wedge dz, dx\wedge dz$ are 2-forms. The $dx\wedge dy\wedge dz$ is a 3-form. 

Differential of arclength is not a differential form$$ds=\sqrt{dx^2+dy^2+dz^2}$$ 

Differential of arclength on a surface is not a differential form$$ds=\sqrt{Edu^2+2Fdudv+Gdv^2}$$
Differential of surface area is not a differential form$$dS=\sqrt{EG-F^2}dudv$$

The variables $x,y,z$ are coordinates, $u,v,t$ are parameters. We have not yet made any specific parameterizations. 

## Exact, Closed

The functions $P,Q$ is a vector field $\vec F=\{P,Q\}$. For example, $P=-\frac{y}{\sqrt{x^2+y^2}}$,  $Q=\frac{x}{\sqrt{x^2+y^2}}$. Compute closedness we can see
$$P_y=\frac{\partial P}{\partial y}=(y^2/(x^2 + y^2) - 1)/\sqrt{(x^2 + y^2)}$$
$$Q_x=\frac{\partial Q}{\partial x}=(1 - x^2/(x^2 + y^2))/\sqrt{(x^2 + y^2)}$$
and
$$P_y-Q_x=[(y^2/(x^2 + y^2) - 1)-(1 - x^2/(x^2 + y^2))]/\sqrt{(x^2 + y^2)}=-1/\sqrt{(x^2+y^2)}$$
So that in this example $P_y \neq Q_x$, the differential form is not closed, there is no $f$ such that total differential $df=Pdx+Qdy$.

```{r d1}
require(Deriv) 
P="-y/sqrt(x^2+y^2)"
Q= "x/sqrt(x^2+y^2)"

Deriv(P,"y") 
Deriv(Q,"x") 

x=3
y=4
#closedness
eval(parse(text=Deriv(P,"y")))-eval(parse(text=Deriv(Q,"x")))
```
Here we just saying an arbitrary $\vec F=\{P,Q\}$ differential form may not be exact, there may not be an $f$ to make $df=Pdx+Qdy$ holds.

**THEOREM 1.4.1** (Donu Arapura 2016) If $F(x, y)dx+G(x, y)dy$ is a closed form on all of $R^2$ with $C^1$ coefficients, then it is exact.

For example, $P=-\frac{y}{x^2+y^2}$,  $Q=\frac{x}{x^2+y^2}$, we just removed square root  from divisors unlike the example above. Note vector $\{x=0,y=0\}$ cannot be a point in this example. So this example cannot be on all of $R^2$, this violates **THEOREM 1.4.1**. Compute closedness we can see 0 as result, it is closed actually. 
```{r d2}
P="-y/(x^2+y^2)"
Q= "x/(x^2+y^2)"

x=3
y=4
#closedness
eval(parse(text=Deriv(P,"y")))-eval(parse(text=Deriv(Q,"x")))
```
One more observation, let $\omega=-\frac{y}{x^2+y^2}dx+\frac{x}{x^2+y^2}dy$. with some reparameterization we find
$$\omega=d\theta$$
$\theta$ is a polar angle, it may not be unique. $d\theta$ can be $2\pi +d\theta, 4\pi+d\theta$ etc. Therefore we do not have the exactness to have $f$ to make $df=Pdx+Qdy$ holds.

## Integral

A differential is just an integral omitted parameter $t$. When we want to make the integration we just put the parameter $t$ back. These are from our studies of differential geometry. The $x,y$ are functions of $t$, $x=x(t),y=y(t)$ is a curve in Eucliean2.

Integral of vector field $P,Q$ with differential form $Pdx+Qdy$ over the $x,y$ curve becomes a line integral, 

**DEFINITION 1.4.2** (Donu Arapura 2016)
$$\int_CPdx+Qdy=\int^b_a [P(x(t),y(t))\frac{dx}{dt}+Q(x(t),y(t))\frac{dy}{dt}]dt$$ 

https://tutorial.math.lamar.edu/Classes/CalcIII/LineIntegralsPtII.aspx

For example, let $P=yx^2$, $Q=\sin(\pi y)$ and curve $\vec r=\{x=t,y=2+2t\}$, evaluate $\int_C Pdx+Qdy$ along $\vec r$, with $0\le t \le 1$.
$$\int^b_a [P(x(t),y(t))\frac{dx}{dt}+Q(x(t),y(t))\frac{dy}{dt}]dt=\\
\int_0^1(2+2t)(t)^2\frac{dt}{dt}dt+\int_0^1\sin(\pi(2+2t))\frac{d(2+2t)}{dt}dt$$
We have just specified the $t$. 

## Green's Theorem

I have worked out few examples from these pages.

https://tutorial.math.lamar.edu/classes/calcIII/GreensTheorem.aspx

https://math.stackexchange.com/questions/1636427/whats-the-meaning-of-c1r

https://www.wolframalpha.com/input?i=double+integral+calculator

Example 1. Evaluate $\oint_{C}xydx+x^2y^3dy$, over area marked by points $\{0,0\}, \{1,0\},\{1,2\}$. 
```{r e1}
require(Deriv)
require(pracma)

P="x*y"
Q="x^2*y^3"

cross=paste0("(",Deriv(Q,"x"),")-(",Deriv(P,"y"),")")
cross
fun <- function(x,y) {(2 * (x * y^3))-(x)}
integral2(fun, 0, 1, 0, function(x) {2*x} ) #2/3
```

Example 2. Evaluate $\oint_C y^3dx−x^3dy$ where $C$ is the positively oriented circle of radius 2 centered at the origin.

Let's generate points on the circle, and plot vectors on the circle. If we only plot a vector, the vector is rooted at the origin. So we need to move the vectors onto the circle by +x and +y.
```{r e2}
interval=10
t=seq(0,360,by=interval)
t=t/360*2*pi
radius=2
x=radius*cos(t)
y=radius*sin(t)
field=cbind(y^3,-x^3)
#plot(field)

plot(x,y,xlim=c(-15,15),ylim=c(-7.5,7.5))# Draw Base R plot with arrow

arrows(x0 = x,
       y0 = y,
       x1 = (field[,1]+x ),
       y1 = (field[,2]+y ))
```

Only use the LHS of Green's theorem,
$$\oint_C Pdx+Qdy$$
We generate points on the circle, and use diff() command to compute $dx$ and $dy$ differentials. As long as the pieces are small enough we can approximate the differentials. And integral, as we know, is a product sum.
```{r e22}
 
interval=.1
t=seq(0,360,by=interval)
t=t/360*2*pi
radius=2
x=radius*cos(t)
y=radius*sin(t)
field=cbind(y^3,-x^3)


#diff(x)
#diff(y)

field=field[-1,]
sum(field[,1]*diff(x)+field[,2]*diff(y))/pi #24

```

Example 3. Evaluate $\oint_C y^3dx−x^3dy$ where $C$ are the two circles of radius 2 and radius 1 centered at the origin with positive orientation. 

We again only use the LHS of Green's theorem, $\oint_C Pdx+Qdy$, do it on curve not area. Just work out half of it, $45/2/2$, is enough.

Note, the first curve is counter clock wise, the second curve is clock wise. We have examples in class that make a new set parameter and functions to make the curve going in reverse direction. But we have rev() command in R.
```{r e3}
interval=.1
t1=seq(0,180,by=interval)
t1=t1/360*2*pi
t2=rev(t1)
 
x1=2*cos(t1)
y1=2*sin(t1)
field1=cbind(y1^3,-x1^3)

field1=field1[-1,]
c1=sum(field1[,1]*diff(x1)+field1[,2]*diff(y1))/pi

x2=1*cos(t2)
y2=1*sin(t2)
field2=cbind(y2^3,-x2^3)

field2=field2[-1,]
c2=sum(field2[,1]*diff(x2)+field2[,2]*diff(y2))/pi

c1+c2
45/2/2
```

Reference: Introduction to differential forms, Donu Arapura, May 6, 2016 Purdue University
